
---
title: "White Wine Quality Analysis by Priyaranjan JC."
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Required Packages, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
```
Data set contains  white variants of the Portuguese "Vinho Verde" wine.
Number of Instances of white wine - 4898.
Number of Attributes: 11 + output attribute(quality)
```{r Loading the Data Set, echo=FALSE, message=FALSE, warning=FALSE}
wine <- read.csv("wineQualityWhites.csv")
```
**Description for Atrributes and Output Variable :**

* *X :* serial number for instances.

* *fixed acidity: *most acids involved with wine or fixed or nonvolatile (do not evaporate readily)

* *volatile acidity: *the amount of acetic acid in wine, which at too high of levels can lead to an unpleasant, vinegar taste

* *citric acid:* found in small quantities, citric acid can add 'freshness' and flavor to wines

* *residual sugar:* the amount of sugar remaining after fermentation stops, it's rare to find wines with less than 1 gram/liter and wines with greater than 45 grams/liter are considered sweet

* *chlorides:* the amount of salt in the wine

* *free sulfur dioxide:* the free form of SO2 exists in equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial growth and the oxidation of wine

* *total sulfur dioxide:* amount of free and bound forms of S02; in low concentrations, SO2 is mostly undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine

* *density:* the density of water is close to that of water depending on the percent alcohol and sugar content

* *pH:* describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 (very basic); most wines are between 3-4 on the pH scale

* *sulphates:* a wine additive which can contribute to sulfur dioxide gas (S02) levels, wich acts as an antimicrobial and antioxidant

* *alcohol:* the percent alcohol content of the wine


* *quality :* Contains a score between 0 and 10.

# Univariate Plots Section
```{r Name and type of the variables, echo=FALSE, message=FALSE, warning=FALSE}
# Dropping X ,as it is just a serial number and does not hold any valuable information.
wine <- wine[,2:13]
str(wine)
```
```{r Summary of attributes, echo=FALSE, message=FALSE, warning=FALSE}
summary(wine)
```


```{r Correlation between quality and other variables, echo=FALSE, message=FALSE, warning=FALSE}
cor(x = wine, y = wine$quality)
# converting quality as a factor
wine$quality <- factor(wine$quality)
```
#### Wine Quality
```{r Historgram plot of quality, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= quality),data = wine)+geom_bar()+ggtitle(label = "Distribution of Wine Quality")
```

Quality variable is unbalanced.It follows a normal distribution.Wine having quality 6 is the most common wine.

#### Fixed.Acidity
```{r Distribution of fixed.acidity, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= fixed.acidity),data = wine)+
  geom_histogram(binwidth = 0.1)+
  ggtitle(label = "Distribution of fixed.acidity")+
  scale_x_continuous(breaks = seq(3,12,0.5))
```


fixed acidity ranges between 3.8 and 14.2. But most of the values are distributed acrsoo 5.5 to 8.5 range.
Considering only 0.95 quantile data(removing outliers) for visualization.

```{r Distribution of fixed.acidity between 5th and 95th percentile, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= fixed.acidity),data = wine)+
  geom_histogram(binwidth = 0.1)+
  ggtitle(label = "Distribution of fixed.acidity between 5th and 95th percentile")+
  scale_x_continuous(breaks = seq(5,8,0.25))+
  coord_cartesian(xlim=c(quantile(wine$fixed.acidity,probs = 0.05),quantile(wine$fixed.acidity,probs = 0.95)))
```
Fixed acidity property of the wine follows mostly a normal distribution between 5th and 95th percentile of the values.

#### Volatile Acidity

```{r Distribution of volatile.acidity, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= volatile.acidity),data = wine)+
  geom_histogram(binwidth = 0.01)+
  ggtitle(label = "Distribution of volatile.acidity")+
  scale_x_continuous(breaks = seq(0.0800,1.1,0.1))
```
High amounts of volatile.acidity will make the wine taste like vinegar.So,most of the acidity is below 0.48 percent.

```{r Distribution of volatile.acidity with 95 percentile, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= volatile.acidity),data = wine)+
  geom_histogram(binwidth = 0.01)+
  ggtitle(label = "Distribution of volatile.acidity with 95th percentile")+
  scale_x_continuous(breaks = seq(0.0800,1.1,0.05))+
  coord_cartesian(xlim=c(0.08,quantile(wine$volatile.acidity,probs = 0.95)))
```
After removing outliers(95th percent quantile),we can observe that volatile.acidity follows a near normal distribution.
In futher investigation we need to find if acidity above 0.48 will result in low quality wines.

#### Citric Acid

```{r Distribution of citric acid, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= citric.acid),data = wine)+
  geom_histogram(binwidth = 0.01)+
  ggtitle(label = "Distribution of Ctric Acid")+
  scale_x_continuous(breaks = seq(0,1.5,0.1))
```
Citric acid is in very low quantitles when compared to fixed acid percent.It is more in the range of volatile acid range.There are few wines with nearly zero citirc acid percent. Range of citric acid is between 0.00 and 1.66.But most the values reside between 0.2 and 0.5.

```{r Distribution of citric acid between 5th and 95th percentile, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= citric.acid),data = wine)+
  geom_histogram(binwidth = 0.01)+
  ggtitle(label = "Distribution of Ctric Acid between 5th and 95th percentile")+
  scale_x_continuous(breaks = seq(0,1.5,0.05))+
  coord_cartesian(xlim=c(quantile(wine$citric.acid,probs = 0.05),quantile(wine$citric.acid,probs = 0.95)))
```
Between 5th and 95th percentile, citric acid moslty follows normal distribution.

#### Residual sugar
```{r Distribution of Residual Sugar, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= residual.sugar),data = wine)+
  geom_histogram(binwidth = 0.1)+
  ggtitle(label = "Distribution of Residaul Sugar")+
  scale_x_continuous(breaks = seq(1,66,5))
```
Most of the values in Resudual sugar are between 0-20.But there are few very large values.
```{r Distribution of Residual Sugar, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= residual.sugar),data = wine)+
  geom_histogram(binwidth = 0.1)+
  ggtitle(label = "Distribution of Residaul Sugar")+
  scale_x_log10()+
  coord_cartesian(xlim=c(quantile(wine$residual.sugar,probs = 0.05),quantile(wine$residual.sugar,probs = 0.95)))

```


```{r Distribution of Residual Sugar between 5th and 95th percentile, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= residual.sugar),data = wine)+
  geom_histogram(binwidth = 0.1)+
  ggtitle(label = "Distribution of Residaul Sugar between 5th and 9th percentile.")+
  scale_x_continuous(breaks = seq(1,16,1))+
  coord_cartesian(xlim=c(quantile(wine$residual.sugar,probs = 0.05),quantile(wine$residual.sugar,probs = 0.95)))
```
Lot of the wines has residual sugar values of  around 1.5 and residual values are uniformly distributed from 3 to 16.

#### chlorides
```{r Distribution of chlorides, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= chlorides),data = wine)+
  geom_histogram(binwidth = 0.001)+
  ggtitle(label = "Distribution of chlorides")+
  scale_x_continuous(breaks = seq(0.00900,0.34600,0.05))
```

```{r Distribution of chlorides between 5th and 95th percentile, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= chlorides),data = wine)+
  geom_histogram(binwidth = 0.001)+
  ggtitle(label = "Distribution of chlorides between 5th and 95th percentile")+
  scale_x_continuous(breaks = seq(0.00900,0.05900,0.005))+
  coord_cartesian(xlim=c(quantile(wine$chlorides,probs = 0.05),quantile(wine$chlorides,probs = 0.95)))
```
Chlorides between 5th and 95th percentile has normal distribution.

#### Free sulphur dioxide
```{r Distribution of Free sulphur dioxide, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= free.sulfur.dioxide),data = wine)+
  geom_histogram(binwidth = 1)+
  ggtitle(label = "Distribution of Free sulphur dioxide")+
  scale_x_continuous(breaks = seq(0,300,20))
```
Most of free sulfur dioxide content is between rangte 0 and 80. There are few very high values of free.sulfur.dioxide.

```{r Distribution of Free sulphur dioxide between 5th and 95th percentile, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= free.sulfur.dioxide),data = wine)+
  geom_histogram(binwidth = 1)+
  ggtitle(label = "Distribution of Free sulphur dioxide between 5th and 95th percentile")+
  scale_x_continuous(breaks = seq(0,70,5))+
  coord_cartesian(xlim=c(quantile(wine$free.sulfur.dioxide,probs = 0.05),quantile(wine$free.sulfur.dioxide,probs = 0.95)))
```
Between 5th and 95th percentile of free.sulfur.dioxide values follow a normal distribution.
#### Total sulfur dioxide
```{r Distribution of total sulfur dioxide, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= total.sulfur.dioxide),data = wine)+
  geom_histogram(binwidth = 1)+
  ggtitle(label = "Distribution of total sulfur dioxide")+
  scale_x_continuous(breaks = seq(0,400,50))
```
```{r Distribution of total sulfur dioxide between 5th and 95th percentile, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= total.sulfur.dioxide),data = wine)+
  geom_histogram(binwidth = 1)+
  ggtitle(label = "Distribution of total sulfur dioxide between 5th and 95th percentile")+
  scale_x_continuous(breaks = seq(0,210,10))+
  coord_cartesian(xlim=c(quantile(wine$total.sulfur.dioxide,probs = 0.05),quantile(wine$total.sulfur.dioxide,probs = 0.95)))
```


#### Density

```{r Distribution of density, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= density),data = wine)+
  geom_histogram(binwidth = 0.0001)+
  ggtitle(label = "Distribution of density")+
  scale_x_continuous(breaks = seq(0.9871,1.0390,0.005))
```


```{r Distribution of density with 95th percentile, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= density),data = wine)+
  geom_histogram(binwidth = 0.0001)+
  ggtitle(label = "Distribution of density between 5ht and 95th percentile")+
  scale_x_continuous(breaks = seq(0.9871,1.0390,0.005))+
  coord_cartesian(xlim=c(quantile(wine$density,probs = 0.05),quantile(wine$density,probs = 0.95)))
```
After Removing the outliers,Density distribution is little closer to normal distribution.

#### ph
```{r Distribution of ph, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= pH),data = wine)+
  geom_histogram(binwidth = 0.01)+
  ggtitle(label = "Distribution of pH levels")+
  scale_x_continuous(breaks = seq(2.7,3.8,0.1))
```
```{r Distribution of ph between 5th and 95th percentile, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= pH),data = wine)+
  geom_histogram(binwidth = 0.01)+
  ggtitle(label = "Distribution of pH levels between 5th and 95th percentile")+
  scale_x_continuous(breaks = seq(2.7,3.8,0.1))+
  coord_cartesian(xlim=c(quantile(wine$pH,probs = 0.05),quantile(wine$pH,probs = 0.95)))
```
Between 5th and 95th percentile ,pH values follow a normal distribution.

#### sulphates
```{r Distribution of sulphates, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= sulphates),data = wine)+
  geom_histogram(binwidth = 0.01)+
  ggtitle(label = "Distribution of Sulphate values")+
  scale_x_continuous(breaks = seq(0.2,1,0.05))
```
sulphate values lies between 0.2 and 1 values. Most of the suphate values are present between 0.35 to 0.65.
```{r Distribution of sulphates between 5th and 95th percentile, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= sulphates),data = wine)+
  geom_histogram(binwidth = 0.01)+
  ggtitle(label = "Distribution of sulphates values between 5th and 95th percentile")+
  scale_x_continuous(breaks = seq(0.2,1,0.05))+
  coord_cartesian(xlim=c(quantile(wine$sulphates,probs = 0.05),quantile(wine$sulphates,probs = 0.95)))
```
between 5th and 95th percentile ,sulphate values follow  a normal distribution.

#### Alcohol
```{r Distribution of alcohol, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= alcohol),data = wine)+
  geom_histogram(binwidth = 0.1)+
  ggtitle(label = "Distribution of Alcohol")+
  scale_x_continuous(breaks = seq(8,14,0.5))
```
Range of the alcohol is between 8.5 and 14.

```{r Distribution of alcohol between 5th and 95th percentile, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= alcohol),data = wine)+
  geom_histogram(binwidth = 0.1)+
  ggtitle(label = "Distribution of Alcohol between 5th and 95th percentile.")+
  scale_x_continuous(breaks = seq(8,14,0.25))+
  coord_cartesian(xlim=c(quantile(wine$alcohol,probs = 0.05),quantile(wine$alcohol,probs = 0.95)))
```
Most of the values of the alcohol are present between 8.75 and 12.75.




# Univariate Analysis

### What is the structure of your dataset?
There are total of 4898 white wine samples.
There are 10 properties of the white wine which define the quality of the wine.

Quality of the samples form a normal distribution with quality 6 as the most common quality.
alcohol,density and chlorides are the most correlated with the quality.
Most of the wines has the alcohol content of 9.5 and alcohol percent continues to increase till 14% with decreasing count of samples.
Most of the wine has density between 0.9871 and 1.Few wines are above density 1.
chlorides are distributed between in the range of 0.009 and 0.309.But majority of wines are in the range of 0.029 and 0.054.


### What is/are the main feature(s) of interest in your dataset?
For density,chlorides,acidity values are mostly in the normal range,i.e with 2 standard deviation most of the data is covered.Will the data beyond 2 standard deviation correspond to either low or high quality wines?This has to be decided using further analysis.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

