
---
title: "White Wine Quality Analysis by Priyaranjan JC."
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Required Packages, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
```
Data set contains  white variants of the Portuguese "Vinho Verde" wine.
Number of Instances of white wine - 4898.
Number of Attributes: 11 + output attribute(quality)
```{r Loading the Data Set, echo=FALSE, message=FALSE, warning=FALSE}
wine <- read.csv("wineQualityWhites.csv")
```
**Description for Atrributes and Output Variable :**

* *X :* serial number for instances.

* *fixed acidity: *most acids involved with wine or fixed or nonvolatile (do not evaporate readily)

* *volatile acidity: *the amount of acetic acid in wine, which at too high of levels can lead to an unpleasant, vinegar taste

* *citric acid:* found in small quantities, citric acid can add 'freshness' and flavor to wines

* *residual sugar:* the amount of sugar remaining after fermentation stops, it's rare to find wines with less than 1 gram/liter and wines with greater than 45 grams/liter are considered sweet

* *chlorides:* the amount of salt in the wine

* *free sulfur dioxide:* the free form of SO2 exists in equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial growth and the oxidation of wine

* *total sulfur dioxide:* amount of free and bound forms of S02; in low concentrations, SO2 is mostly undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine

* *density:* the density of water is close to that of water depending on the percent alcohol and sugar content

* *pH:* describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 (very basic); most wines are between 3-4 on the pH scale

* *sulphates:* a wine additive which can contribute to sulfur dioxide gas (S02) levels, wich acts as an antimicrobial and antioxidant

* *alcohol:* the percent alcohol content of the wine


* *quality :* Contains a score between 0 and 10.

# Univariate Plots Section
```{r Name and type of the variables, echo=FALSE, message=FALSE, warning=FALSE}
# Dropping X ,as it is just a serial number and does not hold any valuable information.
wine <- wine[,2:13]
str(wine)
```
```{r Summary of attributes, echo=FALSE, message=FALSE, warning=FALSE}
summary(wine)
```


```{r Correlation between quality and other variables, echo=FALSE, message=FALSE, warning=FALSE}
cor(x = wine, y = wine$quality)
# converting quality as a factor
wine$quality <- factor(wine$quality)
```

```{r Historgram plot of quality, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= quality),data = wine)+geom_bar()+ggtitle(label = "Distribution of Wine Quality")
```

Quality variable is unbalanced.It follows a normal distribution.Wine having quality 6 is the most common wine.

```{r Distribution of alcohol, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= alcohol),data = wine)+
  geom_histogram(binwidth = 0.1)+
  ggtitle(label = "Distribution of Alcohol")+
  scale_x_continuous(breaks = seq(8,14,0.5))
```

Distribution of alcohol as above.Range of the alcohol is between 8.5 and 14.

```{r Distribution of volatile.acidity, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= volatile.acidity),data = wine)+
  geom_histogram(binwidth = 0.01)+
  ggtitle(label = "Distribution of volatile.acidity")+
  scale_x_continuous(breaks = seq(0.0800,1.1,0.1))
```

Considering only 0.95 quantile data(removing outliers) for visualization.
High amounts of volatile.acidity will make the wine take like vinegar.So,most of the acidity is below 0.48.

```{r Distribution of volatile.acidity with 95 percentile, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= volatile.acidity),data = wine)+
  geom_histogram(binwidth = 0.01)+
  ggtitle(label = "Distribution of volatile.acidity with 95th percentile")+
  scale_x_continuous(breaks = seq(0.0800,1.1,0.05))+
  coord_cartesian(xlim=c(0.08,quantile(wine$volatile.acidity,probs = 0.95)))
```

After removing outliers(95th percent quantile),we can observe that volatile.acidity follows a near normal distribution.
In futher investigation we need to find if acidity above 0.48 will result in low quality wines.
```{r Distribution of density, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= density),data = wine)+
  geom_histogram(binwidth = 0.0001)+
  ggtitle(label = "Distribution of density")+
  scale_x_continuous(breaks = seq(0.9871,1.0390,0.005))
```


```{r Distribution of density with 95th percentile, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= density),data = wine)+
  geom_histogram(binwidth = 0.0001)+
  ggtitle(label = "Distribution of density with 95th percentile")+
  scale_x_continuous(breaks = seq(0.9871,1.0390,0.005))+
  coord_cartesian(xlim=c(min(wine$density),quantile(wine$density,probs = 0.95)))
```

After Removing the long tail,Density distribution looks more even.
```{r Distribution of chlorides, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= chlorides),data = wine)+
  geom_histogram(binwidth = 0.001)+
  ggtitle(label = "Distribution of chlorides")+
  scale_x_continuous(breaks = seq(0.00900,0.34600,0.05))
```

```{r Distribution of chlorides with 95h percentile, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= chlorides),data = wine)+
  geom_histogram(binwidth = 0.001)+
  ggtitle(label = "Distribution of chlorides with 95th percentile")+
  scale_x_continuous(breaks = seq(0.00900,0.34600,0.05))+
  coord_cartesian(xlim=c(min(wine$chlorides),quantile(wine$chlorides,probs = 0.95)))
```

Few of the data variables contains outlier values,does these values correspond to quality = 3 and quality=9 wines(having low values)? Is there a linear relationship between attribute and target variable(quality)?


# Univariate Analysis
