
---
title: "White Wine Quality Analysis by Priyaranjan JC."
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Required Packages, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(corrplot)
```
This Data set contains  white variants of the Portuguese "Vinho Verde" wine.
Number of Instances of white wine - 4898.
Number of Attributes: 11 + output attribute(quality)
```{r Loading the Data Set, echo=FALSE, message=FALSE, warning=FALSE}
wine <- read.csv("wineQualityWhites.csv")
```
**Description for Atrributes and Output Variable :**

* *X :* serial number for instances.

* *fixed acidity: *most acids involved with wine or fixed or nonvolatile (do not evaporate readily)

* *volatile acidity: *the amount of acetic acid in wine, which at too high of levels can lead to an unpleasant, vinegar taste

* *citric acid:* found in small quantities, citric acid can add 'freshness' and flavor to wines

* *residual sugar:* the amount of sugar remaining after fermentation stops, it's rare to find wines with less than 1 gram/liter and wines with greater than 45 grams/liter are considered sweet.

* *chlorides:* the amount of salt in the wine.

* *free sulfur dioxide:* the free form of SO2 exists in equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial growth and the oxidation of wine

* *total sulfur dioxide:* amount of free and bound forms of S02; in low concentrations, SO2 is mostly undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine

* *density:* the density of water is close to that of water depending on the percent alcohol and sugar content

* *pH:* describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 (very basic); most wines are between 3-4 on the pH scale

* *sulphates:* a wine additive which can contribute to sulfur dioxide gas (S02) levels, which acts as an antimicrobial and antioxidant

* *alcohol:* the percent alcohol content of the wine


* *quality :* Contains a score between 0 and 10.

# Univariate Plots Section
```{r Name and type of the variables, echo=FALSE, message=FALSE, warning=FALSE}
# Dropping X ,as it is just a serial number and does not hold any valuable information.
wine <- wine[,2:13]
# Saving correlation between different varialbes of wine data.
M <- cor(wine)
str(wine)
```
```{r Summary of attributes, echo=FALSE, message=FALSE, warning=FALSE}
summary(wine)
```


```{r Correlation between quality and other variables, echo=FALSE, message=FALSE, warning=FALSE}
# Correlation between other varialbes and quality.
cor(x = wine, y = wine$quality)
```
#### Wine Quality
```{r Historgram plot of quality, echo=FALSE, message=FALSE, warning=FALSE}
# diving quality in to groups or buckets for analysis
wine$quality.group <- with(wine, cut(wine$quality, breaks = c(2,4,7,10),ordered = TRUE))
wine$quality <- factor(wine$quality,ordered = TRUE)
ggplot(aes(x= quality,fill=quality.group),data = wine)+
  geom_bar()+
  ggtitle(label = "Distribution of Wine Quality")
```

From the above graph we can see that quality variable is unbalanced.It follows a normal distribution.
Wine having quality 6 is the most common wine.We can also observe the distribution of the quality groups(low,medium,high),with medium quality(quality between 5 and 7) being the most frequent.Low and high quality wines are less frequent.

#### Fixed.Acidity
```{r Distribution of fixed.acidity, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= fixed.acidity),data = wine)+
  geom_histogram(binwidth = 0.1)+
  ggtitle(label = "Distribution of fixed.acidity")+
  scale_x_continuous(breaks = seq(3,12,0.5))
```

Fixed acidity ranges between 3.8 and 14.2. But most of the values are distributed across 5.5 to 8.5 range.
By removing the extreme values(outliers) on the both tails,we can gain more insight into actual distribution of varialble.
Considering only between  0.02 and 0.98 quantile values of the varialbe for visualization.
```{r Distribution of fixed.acidity between 5th and 95th percentile, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= fixed.acidity),data = wine)+
  geom_histogram(binwidth = 0.1,fill="light blue")+
  ggtitle(label = "Distribution of fixed.acidity between 2ndh and 98th percentile")+
  scale_x_continuous(breaks = seq(4,9,0.25))+
  coord_cartesian(xlim=c(quantile(wine$fixed.acidity,probs = 0.02),quantile(wine$fixed.acidity,probs = 0.98)))
```

From the above graph we can observe that fixed.acidity between 2nd and 98th percentile of values follows an approximate normal distribution.

#### Volatile Acidity

```{r Distribution of volatile.acidity, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= volatile.acidity),data = wine)+
  geom_histogram(binwidth = 0.01)+
  ggtitle(label = "Distribution of volatile.acidity")+
  scale_x_continuous(breaks = seq(0.0800,1.1,0.1))
```

High amounts of volatile.acidity will make the wine taste like vinegar.So,most of the acidity is below 0.48 percent.Few of the values are beyond this.We have to investigate,does this values result in low or high quality wines?

```{r Distribution of volatile.acidity between 2nd and  98th percentile, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= volatile.acidity),data = wine)+
  geom_histogram(binwidth = 0.01,fill="light blue")+
  ggtitle(label = "Distribution of volatile.acidity between 2nd and 95th percentile")+
  scale_x_continuous(breaks = seq(0.0800,0.5,0.025))+
  coord_cartesian(xlim=c(quantile(wine$volatile.acidity,probs = 0.02),quantile(wine$volatile.acidity,probs = 0.95)))
```

After removing extreme values (95th percent quantile) on bothe sides,we can observe that volatile.acidity follows a near normal distribution.We can see that 95th of values fall below 0.48 volatile acidity.

#### Citric Acid

```{r Distribution of citric acid, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= citric.acid),data = wine)+
  geom_histogram(binwidth = 0.01)+
  ggtitle(label = "Distribution of Ctric Acid")+
  scale_x_continuous(breaks = seq(0,1.5,0.1))
```

Citric acid is in very low quantitles when compared to fixed acid percent.It is more in the range of volatile acid range.There are few wines with nearly zero citirc acid percent. Range of citric acid is between 0.00 and 1.66.But most the values reside between 0.2 and 0.5.

```{r Distribution of citric acid between 5th and 95th percentile, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= citric.acid),data = wine)+
  geom_histogram(binwidth = 0.01,fill="light blue")+
  ggtitle(label = "Distribution of Ctric Acid between 5th and 95th percentile")+
  scale_x_continuous(breaks = seq(0,1.5,0.05))+
  coord_cartesian(xlim=c(quantile(wine$citric.acid,probs = 0.05),quantile(wine$citric.acid,probs = 0.95)))
```

Between 5th and 95th percentile, citric acid moslty follows normal distribution except for very high value at 0.50.

#### Residual sugar
```{r Distribution of Residual Sugar, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= residual.sugar),data = wine)+
  geom_histogram(binwidth = 0.1)+
  ggtitle(label = "Distribution of Residaul Sugar")+
  scale_x_continuous(breaks = seq(1,66,5))
```

Most of the values in Resudual sugar are between 0 and 20.But there are few very large values.Lot of the wines has residual sugar values of  around 1.5 and residual values are uniformly distributed from 3 to 16.But distribution of the residual sugar is little different from the distributions we have seen so far.So,applying log scale to x axis for further investigation.
```{r Distribution of Residual Sugar in log scale, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= residual.sugar),data = wine)+
  geom_histogram(binwidth = 0.05,fill="brown")+
  ggtitle(label = "Distribution of Residaul Sugar in log scale")+
  scale_x_log10(breaks = seq(0,66,10))
```

After applying the log scale to x-axis we can clearly observe that,residual.sugar follows a bimodal distribution.

#### chlorides
```{r Distribution of chlorides, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= chlorides),data = wine)+
  geom_histogram(binwidth = 0.001)+
  ggtitle(label = "Distribution of chlorides")+
  scale_x_continuous(breaks = seq(0.00900,0.34600,0.05))
```

From above historgram we can oberseve that most of the chlorides values fall well below 0.06 value.
```{r Distribution of chlorides between 5th and 95th percentile, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= chlorides),data = wine)+
  geom_histogram(binwidth = 0.001,fill="light blue")+
  ggtitle(label = "Distribution of chlorides between 5th and 95th percentile")+
  scale_x_continuous(breaks = seq(0.00900,0.05900,0.005))+
  coord_cartesian(xlim=c(quantile(wine$chlorides,probs = 0.05),quantile(wine$chlorides,probs = 0.95)))
```

After removing extreme values on both tails,Chlorides values between 5th and 95th percentile follow an approximate normal distribution.

#### Free sulphur dioxide
```{r Distribution of Free sulphur dioxide, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= free.sulfur.dioxide),data = wine)+
  geom_histogram(binwidth = 1)+
  ggtitle(label = "Distribution of Free sulphur dioxide")+
  scale_x_continuous(breaks = seq(0,300,20))
```

Most of free sulfur dioxide content is between rangte 0 and 70.Few values of th free sulfur dioxide are a bit high.

```{r Distribution of Free sulphur dioxide between 5th and 95th percentile, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= free.sulfur.dioxide),data = wine)+
  geom_histogram(binwidth = 1,fill="light blue")+
  ggtitle(label = "Distribution of Free sulphur dioxide between 5th and 95th percentile")+
  scale_x_continuous(breaks = seq(0,70,5))+
  coord_cartesian(xlim=c(quantile(wine$free.sulfur.dioxide,probs = 0.05),quantile(wine$free.sulfur.dioxide,probs = 0.95)))
```

As we can oberserve in the above plot,between 5th and 95th percentile of free.sulfur.dioxide values follow a normal distribution.

#### Total sulfur dioxide
```{r Distribution of total sulfur dioxide, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= total.sulfur.dioxide),data = wine)+
  geom_histogram(binwidth = 1)+
  ggtitle(label = "Distribution of total sulfur dioxide")+
  scale_x_continuous(breaks = seq(0,400,50))
```

Total sulfur dioxide also follows a simiar distribution pattern to free sulphur dioxide.Most of th total sulphur dioxide values fall in the range between 75 and 225.

#### Density

```{r Distribution of density, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= density),data = wine)+
  geom_histogram(binwidth = 0.0001)+
  ggtitle(label = "Distribution of density")+
  scale_x_continuous(breaks = seq(0.9871,1.0390,0.005))
```

Density of the wine also follows similar patterns to other attributes of wine.Most of the dessity values fall between 0.99 and 1.

```{r Distribution of density with 95th percentile, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= density),data = wine)+
  geom_histogram(binwidth = 0.0001,fill="light blue")+
  ggtitle(label = "Distribution of density between 5ht and 95th percentile")+
  scale_x_continuous(breaks = seq(0.9871,1.0390,0.005))+
  coord_cartesian(xlim=c(quantile(wine$density,probs = 0.05),quantile(wine$density,probs = 0.95)))
```

After Removing the outliers,Density distribution is little closer to normal distribution.

#### ph
```{r Distribution of ph, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= pH),data = wine)+
  geom_histogram(binwidth = 0.01)+
  ggtitle(label = "Distribution of pH levels")+
  scale_x_continuous(breaks = seq(2.7,3.8,0.1))
```

pH values also as other attributes follow,normal distribution with some extreme values.

#### sulphates
```{r Distribution of sulphates, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= sulphates),data = wine)+
  geom_histogram(binwidth = 0.01)+
  ggtitle(label = "Distribution of Sulphate values")+
  scale_x_continuous(breaks = seq(0.2,1,0.05))
```

sulphate values lies between 0.2 and 1 values. Most of the suphate values are present between 0.35 to 0.65.
Between 5th and 95th percentile ,sulphate values follow  a normal distribution.

#### Alcohol
```{r Distribution of alcohol, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x= alcohol),data = wine)+
  geom_histogram(binwidth = 0.1)+
  ggtitle(label = "Distribution of Alcohol")+
  scale_x_continuous(breaks = seq(8,14,0.5))
```

Distribution of the alcohol is bit different when compared to other attributes.Range of the alcohol is between 8.5 and 14.With most of the values being around 9.0 and following decreasing frequency trends from then on.


# Univariate Analysis

### What is the structure of your dataset?
There are total of 4898 white wine samples.
There are 10 properties of the white wine which define the quality of the wine.

Quality of the samples form a normal distribution with quality 6 as the most common quality.
alcohol,density and chlorides are the most correlated with the quality.
Most of the wines has the alcohol content of 9.5 and alcohol percent continues to increase till 14% with decreasing count of samples.
Most of the wine has density between 0.9871 and 1.Few wines are above density 1.
chlorides are distributed between in the range of 0.009 and 0.309.But majority of wines are in the range of 0.029 and 0.054.


### What is/are the main feature(s) of interest in your dataset?
Almost all of the varialbes(except residual sugar and alcohol) have a similar distribution.For these attributes,Values between 5th and 95th values follows a normal like distribution.All of the varialbes has values beyond 3 standard deviation from mean.
Residual sugar forms a bimodal distribution.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?
Volatile acidity, total sulphur dioxide,residual sugar are directly related to taste of the wine. Improper proportions of these will lead to bad taste of wine. Alcohol,density and chlorides determines forms the important chemical properties of the wine,these properites might contribute a lot to quality of the wine. 

### Did you create any new variables from existing variables in the dataset?
I created a new varialbe called quality.group from quality.Quality of the wines are divided into three groups,quality of the wine between 2 and 4 is considered low quality wine,between 5 and 7 as medium quality wine and between 7 and 10 as high quality wine.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

All of the varialbes has few values beyond 3 standard deviatations from the mean, When  these variables are observed between 5th and 95th values, most of them followed normal distribution.When distribution of the quality of the wine is observed, most of the wines have medium quality and very few has high and low quality.Does the vaues, which are more than 3 standard deviation from the mean of variables,contribute to quality of the wine.Does these values causes the wine to have a high or low quality.This hypothesis has to be further investigated.

# Bivariate Plots Section

Correlation plot between different variables.

```{r Correlation between variables, echo=FALSE, message=FALSE, warning=FALSE}
corrplot.mixed(M,tl.pos = 'd')
```

```{r Correlation between fixed and volatile acidity, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=fixed.acidity,y=volatile.acidity),data=wine)+
  geom_point(alpha = 0.1)+
  geom_smooth()+
  ggtitle(label = "Scatterplot between volatile and fixed acidity")
```

We can see in that points are randomly scattered, except for the dense plots in middle of the plot.
By observing geom smooth plot,We can clearly see there is no strong correlation between Volatile acidity and fixed acidity.

```{r Correlation between density and residual sugar full data, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=residual.sugar,y=density),data=wine)+geom_point(alpha=0.05)+
ggtitle(label = "Scatterplot between density and resiudal sugar")
```

we can  clearly  see a positive correlation between density and residual sugar in above graph.Lets consider the same graph with out extreme values(outliers.)

```{r Correlation between density and residual sugar full data with 95th percentiles, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=residual.sugar,y=density),data=wine)+geom_point(alpha=0.2)+
ggtitle(label = "Scatterplot between density and residual sugar with 95th percentile")+
coord_cartesian(xlim=c(0,quantile(wine$residual.sugar,probs = 0.95)),ylim = c(min(wine$density),quantile(wine$density,probs = 0.95)))+
  geom_smooth()
```

Removing above 95th percentile values from density and residual sugar,shows a strong relationship between density and residual sugar.

```{r Correlation between density and total sulphur dioxide, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=total.sulfur.dioxide,y=density),data=wine)+geom_point(alpha=0.05)+
ggtitle(label = "Scatterplot between density and total sulphate dioxide")
```

Above scatterplot suggest a small positive correlation between dessity and total.sulfur.dioxide.

```{r Correlation between density and total sulphur dioxide with 95th percentile, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=total.sulfur.dioxide,y=density),data=wine)+geom_point(alpha=0.2)+
ggtitle(label = "Scatterplot between density and total sulphur dioxide with 95th percentile")+
coord_cartesian(xlim=c(min(wine$total.sulfur.dioxide),quantile(wine$total.sulfur.dioxide,probs = 0.95)),ylim = c(min(wine$density),quantile(wine$density,probs = 0.95)))+
  geom_smooth()
```

scatter plot between density and total sulphur dioxide with 95th percentile value clearly shows a positve correlation between these two variables.

```{r Correlation between density and alcohol, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=alcohol,y=density),data=wine)+geom_jitter(alpha=0.2)+
ggtitle(label = "Scatterplot between density and alcohol")
```

```{r Correlation between density and alcohol with 95th percentiles, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=alcohol,y=density),data=wine)+geom_jitter(alpha=0.1)+
ggtitle(label = "Scatterplot between density and alcohol with 95th percentile")+
  coord_cartesian(ylim = c(min(wine$density),quantile(wine$density,probs = 0.95)))+
geom_smooth()
```

Above two graphs between alcohol and density clearly shows a negative correlation.

```{r boxplot of density with quality groups, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=quality.group,y=density),data=wine)+geom_boxplot(outlier.colour="red")+
ggtitle(label = "boxplot of density with quality groups")
```

From above box plot , high quality group of wines has less density in most cases when compared to low quality and medium quality groups.


```{r boxplot of alcohol with quality group, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=quality.group,y=alcohol),data=wine)+geom_boxplot(outlier.colour="red")+
ggtitle(label = "boxplot of alcohol with quality group")
```

Above box plot cleary depicts that high quality wines contains high alcohol content.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?
* There is a small negative correlation between density of the wine and quality of the wine,may be because of the strong positive correlation between residual.sugar and density,Wines with high density may tend to  be sweeter,which may not be desired  taste of the wine.
* There is small positive correlation between alcohol and quality of the wine,wines with more alcohol  content are preferred to wines with less alcohol content.


### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?
There is negative correlation between density and alcohol content and positive relation between density and residual sugar.As sugar content increases density increases or if alcohol(poportinal to water content) increases density decreases.
When compared to fixed acidity values,volatile acidity values are prettyy less and does not have great impact on overall pH of the wine. Fixed acidity and pH values have negative correlation.


### What was the strongest relationship you found?
Wines residual sugar and density of the wine has a really strong positive correlaiton ship.There is a +0.84 between these variables.


# Multivariate Analysis

```{r Correlation between density and residual sugar  alcohol division, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=residual.sugar,y=density,color=cut(alcohol,3)),data=wine)+
  geom_point(alpha=0.5)+
  scale_color_brewer(type='qual')+
  ggtitle(label = "Scatterplot between density and resiudal sugar with alcohol buckets")+
  coord_cartesian(xlim=c(0,quantile(wine$residual.sugar,probs = 0.95)),ylim = c(min(wine$density),quantile(wine$density,probs = 0.95)))
```

Above plot clearly depicts the positive correlation between residual sugar and density.But we can clearly observe that alcohol content is clearly less for high density wines and vice versa.

```{r Correlation between density and alcohol with quality group division, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=alcohol,y=density,color=quality.group),data=wine)+
  geom_point(alpha=0.5,size=2)+
  scale_color_brewer(type='qual',palette = 2)+
  ggtitle(label = "Scatterplot between density and alcohol with quality group division")+
  coord_cartesian(xlim=c(0,quantile(wine$residual.sugar,probs = 0.95)),ylim = c(min(wine$density),quantile(wine$density,probs = 0.95)))+
  facet_wrap(~quality.group)
```

Above facet wrap clealy states the strong negative correlation between density and alcohol but we can see the same trend following across all quality groups.

```{r Correlation between chlorides and  total sulphur dioxide  with quality facet wrap, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=chlorides,y=total.sulfur.dioxide,color=(quality.group)),data=wine)+
  geom_point(alpha=0.5)+
  scale_color_brewer(type='qual',palette = 2)+
  ggtitle(label = "Scatterplot between density and resiudal sugar with quality group facet wrap")+
  facet_wrap(~quality.group,nrow = 3)
```

Above graph is a scatterplot of total sulfur dioxide and chlorides which facets around quality group distributions.
We can observe that high quality wines have less chloride values or its just that high quality wines does not have enough samples to spread across.Also,total sulfur dioxide for high quality wines is in range between 100 and 200.


```{r Correlation between density and quality with sugar division, echo=FALSE, message=FALSE, warning=FALSE}
# sugar has bimodal distribution,so dividing sugar around median.
sugar.division.around.median <- cut(wine$residual.sugar,breaks = c(min(wine$residual.sugar),5.2,max(wine$residual.sugar)))
ggplot(aes(x=quality,y=density,color=(sugar.division.around.median)),data=wine)+
  geom_jitter(alpha=0.5,size=2.5)+
  scale_color_brewer(type='qual',palette = 1)+
  ggtitle(label = "Scatterplot between density and quality with sugar division")+
  coord_cartesian(ylim = c(min(wine$density),quantile(wine$density,probs = 0.99)))
```

We observed that sugar has bimodal distribution in univariate analysis.So,sugar is divided into two groups one below median(5.2) and one above it.Abvoe plot clearly shows the relationship between density and residual.sugar,low sugar values lead to low density wines.Also,with increasing quality ,density of the wine is decreasing. 


```{r Correlation between free sulfur dioxide and total sulfur dioxide  with alcohol division, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=free.sulfur.dioxide,y=total.sulfur.dioxide,color=(cut(alcohol,2))),data=wine)+
  geom_point(alpha=0.5)+
  scale_color_brewer(type='qual',palette =6)+
  ggtitle(label = "Scatterplot between free sulfur dioxide and total sulfur dioxide with alcohol division")
```



```{r Correlation between ph and fixed acidity  with quality division, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=wine$pH ,y=wine$fixed.acidity,color=(quality.group)),data=wine)+
  geom_point(alpha=0.1)+
  scale_color_brewer(type='qual',palette =6)+
  ggtitle(label = "Scatterplot between density and resiudal sugar with quality group distinction")+
  geom_smooth(aes(shape=quality.group))+
  facet_wrap(~quality.group,shrink = FALSE)
```